---
title: "RSQLite"
author: "Robert A. Stevens"
date: "Wednesday, May 18, 2017"
output: html_document
---

https://cran.r-project.org/web/packages/RSQLite/vignettes/RSQLite.html

Hadley Wickham

2017-01-08

RSQLite is the easiest way to use a database from R because the package itself contains SQLite; no external software is needed. This vignette will walk you through the basics of using a SQLite database.

RSQLite is a DBI-compatible interface which means you primarily use functions defined in the DBI package, so you should always start by loading DBI, not RSQLite:

```{r, comment=NA}
library(DBI)
```

## Creating a new database

To create a new SQLite database, you simply supply the filename to dbConnect():

```{r, comment=NA}
mydb <- dbConnect(RSQLite::SQLite(), "my-db.sqlite")
dbDisconnect(mydb)
unlink("my-db.sqlite")
```

If you just need a temporary database, use either "" (for an on-disk database) or ":memory:" or "file::memory:" (for a in-memory database). This database will be automatically deleted when you disconnect from it.

```{r, comment=NA}
mydb <- dbConnect(RSQLite::SQLite(), "")
dbDisconnect(mydb)
```

## Loading data

You can easily copy an R data frame into a SQLite database with dbWriteTable():

```{r, comment=NA}
mydb <- dbConnect(RSQLite::SQLite(), "")
dbWriteTable(mydb, "mtcars", mtcars)
dbWriteTable(mydb, "iris", iris)
dbListTables(mydb)
```

## Queries

Issue a query with dbGetQuery():

```{r, comment=NA}
dbGetQuery(mydb, 'SELECT * FROM mtcars LIMIT 5')
```

Not all R variable names are valid SQL variable names, so you may need to escape them with ":

```{r, comment=NA}
dbGetQuery(mydb, 'SELECT * FROM iris WHERE "Sepal.Length" < 4.6')
```

If you need to insert the value from a user into a query, don't use paste()! That makes it easy for a malicious attacker to insert SQL that might damager your database or reveal sensitive information. Instead, use a parameterised query:

```{r, comment=NA}
dbGetQuery(mydb, 'SELECT * FROM iris WHERE "Sepal.Length" < :x', 
  params = list(x = 4.6))
```

This is a little more typing, but much much safer.

## Batched queries

If you run a query and the results don't fit in memory, you can use dbSendQuery(), dbFetch() and dbClearResults() to retrieve the results in batches. By default dbFetch() will retrieve all available rows: use n to set the maximum number of rows to return.

```{r, comment=NA}
rs <- dbSendQuery(mydb, 'SELECT * FROM mtcars')
while (!dbHasCompleted(rs)) {
  df <- dbFetch(rs, n = 10)
  print(nrow(df))
}
dbClearResult(rs)
```

## Multiple parameterised queries

You can use the same approach to run the same parameterised query with different parameters. Call dbBind() to set the parameters:

```{r, comment=NA}
rs <- dbSendQuery(mydb, 'SELECT * FROM iris WHERE "Sepal.Length" < :x')
dbBind(rs, param = list(x = 4.5))
nrow(dbFetch(rs))
dbBind(rs, param = list(x = 4))
nrow(dbFetch(rs))
dbClearResult(rs)
```

You can also pass multiple parameters in one call to dbBind():

```{r, comment=NA}
rs <- dbSendQuery(mydb, 'SELECT * FROM iris WHERE "Sepal.Length" = :x')
dbBind(rs, param = list(x = seq(4, 4.4, by = 0.1)))
nrow(dbFetch(rs))
dbClearResult(rs)
```

## Statements

DBI has new functions dbSendStatement() and dbExecute(), which are the counterparts of dbSendQuery() and dbGetQuery() for SQL statements that do not return a tabular result, such as inserting records into a table, updating a table, or setting engine parameters. It is good practice, although currently not enforced, to use the new functions when you don't expect a result.

```{r, comment=NA}
dbExecute(mydb, 'DELETE FROM iris WHERE "Sepal.Length" < 4')
rs <- dbSendStatement(mydb, 'DELETE FROM iris WHERE "Sepal.Length" < :x')
dbBind(rs, param = list(x = 4.5))
dbGetRowsAffected(rs)
dbClearResult(rs)
```
